FGF- and FGFR-coding gene alterations in urothelial cancers

### 2024-05-16 report/markdown/report.Rmd
Analysis report
Figures for the MSK IMPACT cohort. Analysis summary and methods for the analysis report.

opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      dev = "CairoPNG", 
                      dpi = 600)

@@

@@

_Percentages of somatic mutations, gene deletions and amplification present in at least 5% of cancers samples in all investigated cohorts (`r globals$cohort_labs %>% paste(collapse = ', ')`, note: gene amplification and deletion data were not available for the `r globals$cohort_labs["imvigor"]` cohort) are presented in bar plots. Alterations affecting FGF- and FGFR-coding genes are highlighted with bold font. Numbers of analyzed cancer samples are indicated in the plot legend._

@@

flextable::flextable(rep_tabs$cohorts) %>% 
  set_widths(c(3, rep(3.5, 3), 2.8, 2.8)) %>% 
  footnote(1, 5:6, 
           value = as_paragraph('Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  theme_vanilla

@@

_Mutations in FGFR1, FGFR2, FGFR3, and FGFR4 genes in the `r globals$cohort_labs[c("genie", "tcga")] %>% paste(collapse = ' and ')` cohorts were classified by their effects on the coding sequence. Their frequency was expressed as percentage of cancer samples and presented in stack plots. Numbers of analyzed cancer samples are displayed in the plot captions. _

@@

_Mutations in FGFR1, FGFR2, FGFR3, and FGFR4 genes in the `r globals$cohort_labs[c("genie", "tcga")] %>% paste(collapse = ' and ')` cohorts were classified by variant type. Their frequency was expressed as percentage of cancer samples and presented in stack plots. Numbers of analyzed cancer samples are displayed in the plot captions. SNP: single nucleotide polymorphism, DNP: di-nucleotide polymorphism;  ONP: oligo-nucleotide polymorphism._

@@

_Mutations in FGFR1, FGFR2, FGFR3, and FGFR4 genes in the `r globals$cohort_labs[c("genie", "tcga")] %>% paste(collapse = ' and ')` cohorts were classified by protein domain. Their frequency was expressed as percentage of cancer samples and presented in stack plots. Numbers of analyzed cancer samples are displayed in the plot captions._

@@

```{r fig-fgfr1-position, fig.width = figur::convert(rep_figs$fgfr1_position, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr1_position, to = 'in')$h, fig.cap = 'Positions of FGFR1 protein affected by somatic mutations.'}

@@

__Figure \@ref(fig:fig-fgfr1-position). Positions of FGFR1 protein affected by somatic mutations.__

_Somatic mutations in the FGFR1 gene in the `r globals$cohort_labs[c("genie", "tcga")] %>% paste(collapse = ' and ')` were classified by their effects on nucleotide sequence and mapped to amino acids of the FGFR1 protein. Mutation localization and type are visualized in scatter plots below the FGFR1 protein domain schemes. Each point represents a single somatic mutation. Point color codes for protein domain. Total mutation numbers are displayed in the plot captions._

@@

```{r fig-fgfr2-position, fig.width = figur::convert(rep_figs$fgfr2_position, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr2_position, to = 'in')$h, fig.cap = 'Positions of FGFR2 protein affected by somatic mutations.'}

@@

__Figure \@ref(fig:fig-fgfr2-position). Positions of FGFR2 protein affected by somatic mutations.__

_Somatic mutations in the FGFR2 gene in the `r globals$cohort_labs[c("genie", "tcga")] %>% paste(collapse = ' and ')` were classified by their effects on nucleotide sequence and mapped to amino acids of the FGFR2 protein. Mutation localization and type are visualized in scatter plots below the FGFR2 protein domain schemes. Each point represents a single somatic mutation. Point color codes for protein domain. Total mutation numbers are displayed in the plot captions._

@@

```{r fig-fgfr3-position, fig.width = figur::convert(rep_figs$fgfr3_position, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr3_position, to = 'in')$h, fig.cap = 'Positions of FGFR3 protein affected by somatic mutations.'}

@@

__Figure \@ref(fig:fig-fgfr3-position). Positions of FGFR3 protein affected by somatic mutations.__

_Somatic mutations in the FGFR3 gene in the `r globals$cohort_labs[c("genie", "tcga")] %>% paste(collapse = ' and ')` were classified by their effects on nucleotide sequence and mapped to amino acids of the FGFR3 protein. Mutation localization and type are visualized in scatter plots below the FGFR3 protein domain schemes. Each point represents a single somatic mutation. Point color codes for protein domain. Total mutation numbers are displayed in the plot captions._

@@

```{r fig-fgfr4-position, fig.width = figur::convert(rep_figs$fgfr4_position, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr4_position, to = 'in')$h, fig.cap = 'Positions of FGFR4 protein affected by somatic mutations.'}

@@

__Figure \@ref(fig:fig-fgfr4-position). Positions of FGFR4 protein affected by somatic mutations.__

_Somatic mutations in the FGFR4 gene in the `r globals$cohort_labs[c("genie", "tcga")] %>% paste(collapse = ' and ')` were classified by their effects on nucleotide sequence and mapped to amino acids of the FGFR4 protein. Mutation localization and type are visualized in scatter plots below the FGFR4 protein domain schemes. Each point represents a single somatic mutation. Point color codes for protein domain. Total mutation numbers are displayed in the plot captions._

@@

```{r fig-genetic-overlap, fig.width = figur::convert(rep_figs$genetic_overlap, to = 'in')$w, fig.height = figur::convert(rep_figs$genetic_overlap, to = 'in')$h, fig.cap = 'Co-occurrence of the most common genetic alterations in urothelial cancer.'}

rep_figs$genetic_overlap$plot

@@

__Figure \@ref(fig:fig-genetic-overlap). Co-occurrence of the most common genetic alterations in urothelial cancer.__

_Co-occurrence of the most common genetic alterations in urothelial cancers (at least 5% of cancer samples) was investigated in the `r globals$cohort_labs[c("genie", "tcga")] %>% paste(collapse = ' and ')` cohorts by pairwise Jaccard's similarity coefficients and two-dimensional MDS (multi-dimensional scaling) of pairwise Jaccard's distances._

_(A) MDS results are visualized in scatter plots with single genetic alterations depicted as points. Point color codes for alteration type (somatic mutation, deletion, amplification), point size represents frequency of the alteration in the data set. Note that genetic features displayed in the plots close to each other are expected to co-occur in a high fraction of cancer samples. Total numbers of samples are indicated in the plot captions._

_(B) Frequencies of samples with and without genetic alterations are visualized as heat maps of contingency tables for selected pairs of genetic features. Sample percentages and counts within the total sample sets are indicated in the heat map tiles. Jaccard's similarity coefficients and with their false discovery rate-corrected significance ($pFDR(J > 0)$) are presented in the plot captions._

@@

```{r fig-fgfr-overlap, fig.width = figur::convert(rep_figs$fgfr_overlap, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr_overlap, to = 'in')$h, fig.cap = 'Co-occurrence of genetic alterations of FGF and FGFR genes in urothelial cancer.'}

rep_figs$fgfr_overlap$plot

@@

__Figure \@ref(fig:fig-fgfr-overlap). Co-occurrence of genetic alterations of FGF and FGFR genes in urothelial cancer.__

_Co-occurrence of somatic mutations, deletions, and amplifications of FGF- and FGFR-coding genes was investigated in the `r globals$cohort_labs[c("genie", "tcga")] %>% paste(collapse = ' and ')` cohorts by pairwise Jaccard's similarity coefficients and two-dimensional MDS (multi-dimensional scaling) of pairwise Jaccard's distances._

_(A) MDS results are visualized in scatter plots with single genetic alterations depicted as points. Point color codes for alteration type (somatic mutation, deletion, amplification), point size represents frequency of the alteration in the data set. Note that genetic features displayed in the plots close to each other are expected to co-occur in a high fraction of cancer samples. Total numbers of samples are indicated in the plot captions._

_(B) Frequencies of samples with and without genetic alterations are visualized as heat maps of contingency tables for selected pairs of genetic features. Sample percentages and counts within the total sample sets are indicated in the heat map tiles. Jaccard's similarity coefficients and with their false discovery rate-corrected significance ($pFDR(J > 0)$) are presented in the plot captions._

>>>>

opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      dev = "cairo_pdf", 
                      dpi = 600)

@@

# Analysis summary

* Fibroblast growth factor receptor (FGFR) signaling has emerged as a promising therapy target in urothelial cancers with approval of the pan-FGFR inhibitor erdafitinib. However, effects of genetic alterations and modulation of expression of FGF- and FGFR-coding genes in urothelial carcinoma on tumor biology, therapy response, and clinical outcome are still not fully understood.

* The primary goal of the current project was to characterize frequency, co-occurrence, and clinical implications of genetic alterations in FGF- and FGFR-coding genes in large publicly available cohorts of urothelial cancers. The secondary goal was an attempt to establish subsets of urothelial cancer specimens based on differences in their profiles of somatic mutations and gene copy number variants.

* Four collectives of urothelial cancer patients were evaluated: the GENIE BLCA cohort, an urothelial carcinoma subset of the large GENIE project (n = `r nrow(genie$clinic)`) [@Sweeney2017], the predominantly muscle-invasive MSK IMPACT cohort (`r nrow(msk$clinic)`) [@Clinton2022], the muscle-invasive TCGA BLCA collective (n = `r nrow(tcga$clinic)`) [@Liu2018; @Robertson2017], and the IMvigor cohort of late stage, platinum-refractory cancers subjected to an anti-PD-L1 treatment (`r nrow(imvigor$clinic)`) [@Balar2017]. Detailed information on somatic mutations and their effects on protein sequence was provided for the GENIE BLCA, MSK IMPACT, and TCGA BLCA cohorts. Binarized indexes of presence/absence of somatic mutations in selected genes were available for the IMvigor cohort. Binarized data on gene amplifications and deletions were obtained for the GENIE BLCA, MSK IMPACT, and TCGA BLCA collectives. Gene expression quantification by RNA sequencing and assignment to consensus molecular classes [@Kamoun2020] were retrieved for the TCGA BLCA and IMvigor cohorts. Overall survival information was available for the MSK IMPACT, TCGA BLCA, and IMvigor cohorts; data on disease-specific and relapse-free survival was provided for the TCGA BLCA collective. Approximate overall survival was calculated for the GENIE BLCA project participants based on age at enrollment in years and follow-up time in days. Characteristic of the investigated collectives is presented in __Table \@ref(tab:tab-cohorts)__.

* Somatic mutations of _FGFR3_ belonged to the most common genetic features of urothelial cancers and were found in `r expl_genet$stats$mutation %>% filter(variable == 'FGFR3') %>% .$percent %>% range %>% signif(2) %>% paste(collapse = ' to ')`% of cancer samples. Somatic mutations of the remaining _FGFR1_, _FGFR2_, and _FGFR4_ receptor genes were less frequent (`r expl_genet$stats$mutation %>% filter(variable %in% c('FGFR1', 'FGFR2', 'FGFR4')) %>% .$percent %>% range %>% signif(2) %>% paste(collapse = ' to ')`% of cancer samples). Amplifications of the _FGF3_, _FGF4_, and _FGF19_ genes in urothelial cancers result from multiplication of the 11q13 chromosome region, which codes also for an essential oncogene _CCDN1_. Amplifications of _FGF3_, _FGF4_, and _FGF19_ were detected in `r expl_genet$stats$amplification %>% filter(variable %in% c('FGF3', 'FGF4', 'FGF19')) %>% .$percent %>% range %>% signif(2) %>% paste(collapse = ' to ')`% of specimens (__Figures \@ref(fig:fig-general-freq)__ and __\@ref(fig:fig-fgfr-freq)__, __Tables \@ref(tab:tab-genetics)__ and __\@ref(tab:tab-fgf-genetics)__).

* A great majority of somatic mutations of _FGFR1_, _FGFR2_, _FGFR3_, and _FGFR4_ was classified as missense mutations caused by single nucleotide polymorphisms. Particularly for _FGFR3_, they affected protein residues were frequently located apart from the Ig-like domains responsible for ligand biding and besides the kinase domains vital for activation of downstream signaling (__Figures \@ref(fig:fig-mutation-type)__ - __\@ref(fig:fig-protein-domain-split)__, __Table \@ref(tab:tab-fgfr-class)__). A more detailed analysis revealed the Ig-like domain 1 and kinase domain as the mutation hot-spots of _FGFR1_. The kinase domain was the prime target of mutations of _FGFR2_. Hinge regions between the Ig-like domains 1 and 2, as well as residues around the membrane-spanning region between the Ig-like domain 3 and the kinase domain were identified as mutational hot-spots of _FGFR3_. In case of _FGFR4_, the Ig-like domain 1, the hinge region between the Ig-like domain 3 and kinase domain, as well as the kinase domain were prime targets of mutations of _FGFR4_ (__Figures \@ref(fig:fig-fgfr1-position)__ - __\@ref(fig:fig-fgfr4-position)__). It can be hence speculated, that somatic mutations of FGF receptors may affect both the kinase activity and the conformation of the protein resulting in a higher affinity towards ligands or ligand-independent signaling.

* Co-occurrence and exclusivity of genetic features was investigated by analysis of pairwise Jacquard similarity coefficients for pairs of genetic alterations in urothelial cancer samples. Such analysis revealed little overlap between somatic mutations of _FGFR3_ and other common alterations such as _TP53_ and _RB1_ mutations, amplification of the 11q13 chromosome region hallmarked by copy number variants of _FGF3/4/19_ and _CCND1_, as well as deletion of the 9p21 chromosome region with the _MTAP_, _CDKN2A_, and _CDKN2A_ genes (__Figures \@ref(fig:fig-genetic-mds)__ - __Figure \@ref(fig:fig-fgfr-cont-hm)__).

* Concerning co-regulation of expression of FGF- and FGFR-coding genes, _FGFR1_ and _FGFR3_ mRNA levels were found to be weakly but significantly negatively associated. Increased expression of _FGFR1_ in the cancer tissue correlated significantly with upregulation of genes coding for multiple FGF ligands (_FGF1_, _FGF2_, _FGF5_, _FGF7_, _FGF9_, _FGF10_, and _FGF18_). This phenomenon was absent for _FGFR3_ (__Figure \@ref(fig:fig-fgfr-correlation)__). In a differential gene expression analysis of cancers with and without _FGFR3_ mutations, we observed that presence of _FGFR3_ mutations translated in significantly and strongly increased _FGFR3_ mRNA levels. By contrast, cancers with _FGFR3_ mutations had significantly lower expression of _FGF2_, _FGF5_, and _FGF7_ (__Figure \@ref(fig:fig-fgfr-dge)__, __Table \@ref(tab:tab-fgfr3-dge)__). This suggests two mechanisms leading to activation of pro-oncogenic FGFR signaling in urothelial carcinoma. In _FGFR3_ mutant tumors, FGFR signaling may be triggered by over-expression of FGFR3 protein, improved FGF-affinity, or even in an ligand-independent manner. Alternatively, concomitant over-expression of _FGFR1_ with its ligands may activated FGFR signaling in auto- or paracrine manner.

* No consistent differences in total mutation burdens (TMB), counts of protein-affecting somatic mutations, or counts of gene deletions were found for urothelial cancers with and without _FGFR3_ mutations. Gene amplification numbers were in turn significantly lower in _FGFR3_ mutated neoplasms as compared with _FGF3_ wild-type tumors. Effect size of this difference was small (__Figure \@ref(fig:fig-fgfr-tmb)__).

* There were no significant differences in age and gender distribution between patients with wild-type and _FGFR3_-mutated cancers. _FGFR3_ mutant cancers tended towards lower pathological stages, this effect was significant in the MSK IMPACT cohort but not in the TCGA BLCA collective (__Table \@ref(tab:tab-fgfr3-clinic)__). In the GENIE BLCA and MSK IMPACT cohorts, patients with _FGFR3_ mutated tumors demonstrated improved overall survival as compared with bearers of WT neoplasms. This phenomenon was also evident in the TCGA BLCA collective, it was only statistically significant upon exclusion of pT4 cancer patients (entire cohort: p = `r signif(fgfr_os$test$tcga$p_adjusted, 2)`, without pT4 cancers: p = `r signif(fgfr_os$test$tcga_wo_t4$p_adjusted, 2)`). There were no significant differences in disease-specific or relapse-free survival between the _FGFR3_ mutation strata in the TCGA BLCA cohort, which may however result from insufficient number of events of interest (tumor-related deaths and relapses, __Figures \@ref(fig:fig-fgfr-os)__ and __Figure \@ref(fig:fig-fgfr-rfs)__).

* Analysis of effects of amplification of the 11q13 chromosome region on expression of FGF- and FGFR-coding genes was possible solely for the TCGA BLCA collective. Presence of this genetic aberration resulting in amplification of the _FGF3_, _FGF4_, _FGF19_, and _CCND1_ genes translated into lower expression of _FGFR1_ and significantly higher levels of _FGF3_, _FGF4_, and _FGF19_ mRNA (__Figure \@ref(fig:fig-amp11q13)__). However, effects of those differences was weak and, in case of _FGF3_ and _FGF4_, the expression levels were extremely low as compared with other FGF ligand-coding genes. Hence, biological relevance of the 11q13 amplification for FGFR signaling remains unclear.

* Basal/squamous-like cancers followed by luminal papillary tumors constituted the largest consensus molecular classes [@Kamoun2020] of urothelial cancers in the TCGA BLCA and IMvigor cohorts (__Figure \@ref(fig:fig-fgfr-subtypes)A__). in both cohorts, _FGFR3_ mutated cancers were found to be significantly enriched in the papillary luminal class (TCGA: `r sub_genet$stats$tcga %>% filter(variable == 'FGFR3_mutation', consensusClass == 'LumP') %>% .$percent %>% signif(2)`%, IMvigor: `r sub_genet$stats$imvigor %>% filter(variable == 'FGFR3_mutation', consensusClass == 'LumP') %>% .$percent %>% signif(2)`% of specimens). In the TCGA BLCA cohort, amplifications of _FGF3_, _FGF4_, and _FGF19_ genes resulting from multiplication of the 11q13 chromosome region were found in luminal papillary (`r sub_genet$stats$tcga %>% filter(variable %in% c('FGF3_amplification', 'FGF4_amplification', 'FGF19_amplification'), consensusClass == 'LumP') %>% .$percent %>% range %>% signif(2) %>% paste(collapse = ' to ')`%), stroma-rich (`r sub_genet$stats$tcga %>% filter(variable %in% c('FGF3_amplification', 'FGF4_amplification', 'FGF19_amplification'), consensusClass == 'Stroma-rich') %>% .$percent %>% range %>% signif(2) %>% paste(collapse = ' to ')`%), and basal/squamous-like cancers (`r sub_genet$stats$tcga %>% filter(variable %in% c('FGF3_amplification', 'FGF4_amplification', 'FGF19_amplification'), consensusClass == 'Ba/Sq') %>% .$percent %>% .[1] %>% signif(2)`% of specimens, __Figures \@ref(fig:fig-fgfr-subtypes)__ and __\@ref(fig:fig-fgfr-subdetails)__, __Table \@ref(tab:tab-cons-genetics)__).

* mRNA levels of `r sub_dge$anova_significant %>% reduce(intersect) %>% map_chr(~paste0('_', .x, '_')) %>% collapse_and` were found to vary significant between the consensus molecular classes of urothelial carcinomas in both the TCGA BLCA and IMvigor data set (__Table \@ref(tab:tab-cons-dge)__). Expression of _FGFR1_ was the highest in stroma-rich and luminal non-specified tumors and the lowest in luminal papillary cancers. In turn, maximal mRNA levels of _FGFR3_ were detected in luminal papillary cancers and the lowest expression was observed in basal/squamous- and neuroendocrine-like tumors. Effect size of those differences was large (__Figure \@ref(fig:fig-sub-fgfr-expression)__). Expression of all significantly regulated _FGF_ ligands was the lowest in luminal papillary cancers. _FGF2_, _FGF7_, _FGF10_, and _FGF18_ transcript amounts followed the expression pattern of _FGFR1_ and were detected at the highest levels in stroma-rich and non-specified luminal cancers. _FGF5_ was expressed at the highest levels by basal/squamous-like, neuroendocrine-like and stroma-rich neoplasms. Effect sizes of differential regulation of these FGF-coding genes was large (__Figure \@ref(fig:fig-sub-fgf-expression1)__ and __\@ref(fig:fig-sub-fgf-expression2)__). Collectively, the differential gene expression analysis results support the model of activation of FGFR signaling propose above. While luminal papillary cancers may be driven by _FGFR3_ mutations and over-expression resulting in FGFR signaling with little dependence on the ligands, stroma-rich and non-specified luminal cancers may exploit auto- and paracrine signaling fueled by over-expression of _FGFR1_ and its ligands.

* By latent class analysis (LCA, called also latent class modeling), we could establish a classification scheme of urothelial cancers based on presence/absence of the most frequent genetic alterations detected in at least 5% of cancer specimens. This classification scheme with six 'genetic subsets' was developed in the GENIE BLCA training cohort and technically validated in the MSK IMPACT and TCGA BLCA cohort (__Figures \@ref(fig:fig-lca-devel)__ and __\@ref(fig:fig-lca-size)__). The 'mutRB1' subset (`r lca_eval$n_numbers %>% map(filter, clust_id == 'mutRB1') %>% map_dbl(~.x$percent) %>% range %>% signif(2) %>% paste(collapse = ' to ')`% of specimens) was characterized by highly frequent _TP53_, _RB1_ and _ARID1A_ mutations, _E2F3_ amplifications, and nearly absent _FGFR3_ mutations, 11q13 amplification, and 9p21 deletions. The 'oligoMut' subset was hallmarked by frequent mutations of _TP53_ and the lowest overall mutation count. The key features of 'hyperMut' subset were the highest total mutation count, as well as frequent somatic mutations of _ARID1A_, _FGFR3_, _CREBBP_, _ERBB2_, _FBXW7_, and lacking copy number variants of 11q13 and 9p21. The 'del9p21' subset was characterized by relatively common deletions of the chromosome region 9p21, amplification of the 11q13 chromosome region, and highly frequent mutations of _TP53_ and _KDM6A_. In the 'mutFGFR3' subset, mutations of _FGFR3_ and _KDM6A_, and deletions of 9p21 were common, while mutations in _TP53_, _RB1_, and _ERBB2_, and gene amplifications were virtually absent. Finally, the 'ampMDM2' subset was characterized by the highest percentage of _MDM2_, highly common copy number variants of the 11p13 and 9p21 chromosome regions and essentially lacking mutations of _TP53_ and _RB1_ (__Figures \@ref(fig:fig-lca-genet1)__ - __\@ref(fig:fig-lca-details-cna2)__, __Table \@ref(tab:tab-lca-genetics)__).

* Numbers of mutations differed significantly between the genetic subsets with large effect size and were the highest in hyperMut cancers and the lowest in oligoMut and mutFGFR3 tumors. Gene deletion numbers were significantly higher in del9p21 and ampMDM2 than in the remaining genetic subsets in the GENIE BLCA and MSK IMPACT collectives (large effect size). Amplification counts were the highest in del9p21, ampMDM2, and mutRB1 subsets and the lowest in mutFGFR3 cancers; those differences were large and significant in all collectives (__Figures \@ref(fig:fig-lca-tmb)__ and __\@ref(fig:fig-lca-cna-count)__).

* We could observe significant differences in expression of _FGFR1_, _FGFR3_, and _FGFR3_, as well as _FGF2_, _FGF5_, FGF7_, and _FGF17_ genes between the genetic subsets of the TCGA BLCA cohort (__Table \@ref(tab:tab-lca-dge)__). Interestingly, mutFGFR3 cancers were found to over-express _FGFR3_ and, at the same time, to suppress _FGF2_, _FGF5_, and _FGF7_. The highest expression levels of _FGFR1_ were observed in mutRB1 and ampMDM2 genetic subsets; _FGFR2_ was found to be expressed in oligoMut, del9p21, and ampMDM2 at the highest levels (__Figures \@ref(fig:fig-lca-dge-fgfr)__ and __\@ref(fig:fig-lca-dge-fgf)__).

* There were no significant differences in age and gender of cancer patients between the genetic clusters that were shared by the GENIE BLCA, MSK IMPACT, and TCGA BLCA collectives (__Figure \@ref(fig:fig-lca-clinic)__). Significantly lower pathological staging was observed in mutFGFR3 subset in the MSK IMPACT cohort, but this difference could not be corroborated in the TCGA BLCA collective. In the TCGA BLCA data set, mutFGFR3 and ampMDM2 genetic subsets comprised predominantly cancers assigned to the luminal papillary consensus class. Basal/squamous-like consensus class cancers predominated in mutRB1, oligoMut, hyperMut, and del9p21 genetic subsets (__Figure \@ref(fig:fig-lca-stage-class)__). Notably, consensus molecular classes of muscle-invasive urothelial carcinomas were established with bulk cancer transcriptome [@Kamoun2020; @Robertson2017]. The disparity between the molecular consensus classes and the genetic subsets proposed by us suggests that phenotypically similar cancers may exploit highly dissimilar sets of driver genetic alterations and arise from unrelated evolutionary processes.

* Significant differences in overall survival were observed between the genetic subsets in the GENIE BLCA and MSK IMPACT cohorts with the worst survival in mutRB1 subset and favorable prognosis inferred for mutFGFR3 and hyperMut cancers. No significant differences in overall survival were found in the TCGA cohort, which may, however, attributed to insufficient number of deaths in the genetic subsets (__Figure \@ref(fig:fig-lca-os)__). As investigated by univariable Cox regression, patients assigned to hyperMut subset were at roughly 50 - 75% significantly lower death risk, and patients in mutFGFR3 were at approximately 43 - 45% significantly lower death risk as compared with mutRB1 subset in both the GENIE BLCA and MSK IMPACT cohort. The general effect of genetic cluster assignment was found to be independent from age, sex, and pathological tumor stage (likelihood ratio test GENIE BLCA: `r lca_cox$lrt$plot_lab[[1]]`, MSK IMPACT: `r lca_cox$lrt$plot_lab[[2]]`). At the levels of single genetic subsets, hyperMut subset as favorable survival marker remained significant upon adjustment for age, sex, and pathological tumor stage. mutFGFR3 subset assignment was found independent from age and gender in the GENIE BLCA cohort but was confounded by age, sex, and pathological cancer stage in the MSK IMPACT collective (__Figure \@ref(fig:fig-lca-cox)__). We could not discern significant differences in disease-specific and relapse-free survival between the genetic subsets of the TCGA BLCA cohort (not shown).

* Collectively, our analysis results suggest at least two mechanisms of FGFR signaling activation in urothelial cancers: the one exploiting FGFR3 mutations and largely independent on FGF ligands and operative predominantly in papillary cancers, and the other relying on over-expression of FGFR1 and FGF ligands predominantly by stroma-rich and non-specified luminal tumors. Furthermore, we put forward a relatively simple classification scheme of urothelial cancers based on the most common somatic mutations, gene and chromosome region deletions and amplifications established in the large GENIE BLCA collective. The key genetic features of those six genetic clusters such as subset-defining mutations and copy number variants, mutation and variant numbers could be reproduced in the MSK IMAPST and TCGA BLCA, i.e. two independent validation cohorts. Of clinical importance, the genetic subset classification bears consequences for overall survival in a age-, gender-, and stage-independent manner.

* Our analysis bears limitations. Incompleteness of clinical information in the GENIE BLCA cohort precluded an in-depth characteristic of the FGFR3 mutation strata and genetic subsets. Unavailability of exact overall survival in this collective may have interfered with results of significant analyses. Detailed mutation data as well as copy number information were not provided for the IMvigor collective, which made validation of the genetic subsets impossible in this cohort of immunotherapy patients. Finally, gene expression data were not available for the largest GENIE BLCA and MSK IMPACT cohorts, which precluded detailed analyses of differential gene expression, biological phenotype, and consensus molecular classification based on transcriptome.

@@

_Percentages of somatic mutations, gene deletions and amplification present in at least 5% of cancers samples in all investigated cohorts (`r globals$cohort_labs %>% collapse_and`, note: gene amplification and deletion data were not available for the `r globals$cohort_labs["imvigor"]` cohort) are presented in bar plots. Alterations affecting FGF- and FGFR-coding genes are highlighted with bold font. Numbers of analyzed cancer samples are indicated in the plot legend._

@@

flextable::flextable(rep_tabs$cohorts) %>% 
  set_widths(c(3, rep(3, 4), 2.6, 2.6)) %>% 
  footnote(1, 6:7, 
           value = as_paragraph('Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  theme_vanilla

@@

_Mutations in FGFR1, FGFR2, FGFR3, and FGFR4 genes in the `r globals$cohort_labs[c("genie", "msk", "tcga")] %>% collapse_and()` cohorts were classified by their effects on the coding sequence. Their frequency was expressed as percentage of cancer samples and presented in stack plots. Numbers of analyzed cancer samples are displayed in the plot captions. _

@@

_Mutations in FGFR1, FGFR2, FGFR3, and FGFR4 genes in the `r globals$cohort_labs[c("genie", "msk", "tcga")] %>% collapse_and` cohorts were classified by variant type. Their frequency was expressed as percentage of cancer samples and presented in stack plots. Numbers of analyzed cancer samples are displayed in the plot captions. SNP: single nucleotide polymorphism, DNP: di-nucleotide polymorphism;  ONP: oligo-nucleotide polymorphism._

@@

_Mutations in FGFR1, FGFR2, FGFR3, and FGFR4 genes in the `r globals$cohort_labs[c("genie", "msk", "tcga")] %>% collapse_and` cohorts were classified by protein domain. Their frequency was expressed as percentage of cancer samples and presented in stack plots. Numbers of analyzed cancer samples are displayed in the plot captions._

@@

```{r fig-fgfr1-position, fig.width = figur::convert(rep_figs$fgfr1_position, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr1_position, to = 'in')$h, fig.cap = 'Residues of FGFR1 protein affected by somatic mutations.'}

@@

__Figure \@ref(fig:fig-fgfr1-position). Residues of FGFR1 protein affected by somatic mutations.__

_Somatic mutations in the FGFR1 gene in the `r globals$cohort_labs[c("genie", "msk", "tcga")] %>% collapse_and` were classified by their effects on nucleotide sequence and mapped to amino acids of the FGFR1 protein. Mutation localization and type are visualized in scatter plots below the FGFR1 protein domain schemes. Each point represents a single somatic mutation. Point color codes for protein domain. Total mutation numbers are displayed in the plot captions._

@@

```{r fig-fgfr2-position, fig.width = figur::convert(rep_figs$fgfr2_position, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr2_position, to = 'in')$h, fig.cap = 'Residues of FGFR2 protein affected by somatic mutations.'}

@@

__Figure \@ref(fig:fig-fgfr2-position). Residues of FGFR2 protein affected by somatic mutations.__

_Somatic mutations in the FGFR2 gene in the `r globals$cohort_labs[c("genie", "msk", "tcga")] %>% collapse_and` were classified by their effects on nucleotide sequence and mapped to amino acids of the FGFR2 protein. Mutation localization and type are visualized in scatter plots below the FGFR2 protein domain schemes. Each point represents a single somatic mutation. Point color codes for protein domain. Total mutation numbers are displayed in the plot captions._

@@

```{r fig-fgfr3-position, fig.width = figur::convert(rep_figs$fgfr3_position, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr3_position, to = 'in')$h, fig.cap = 'Residues of FGFR3 protein affected by somatic mutations.'}

@@

__Figure \@ref(fig:fig-fgfr3-position). Residues of FGFR3 protein affected by somatic mutations.__

_Somatic mutations in the FGFR3 gene in the `r globals$cohort_labs[c("genie", "msk", "tcga")] %>% collapse_and` were classified by their effects on nucleotide sequence and mapped to amino acids of the FGFR3 protein. Mutation localization and type are visualized in scatter plots below the FGFR3 protein domain schemes. Each point represents a single somatic mutation. Point color codes for protein domain. Total mutation numbers are displayed in the plot captions._

@@

```{r fig-fgfr4-position, fig.width = figur::convert(rep_figs$fgfr4_position, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr4_position, to = 'in')$h, fig.cap = 'Residues of FGFR4 protein affected by somatic mutations.'}

@@

__Figure \@ref(fig:fig-fgfr4-position). Residues of FGFR4 protein affected by somatic mutations.__

_Somatic mutations in the FGFR4 gene in the `r globals$cohort_labs[c("genie", "msk", "tcga")] %>% collapse_and` were classified by their effects on nucleotide sequence and mapped to amino acids of the FGFR4 protein. Mutation localization and type are visualized in scatter plots below the FGFR4 protein domain schemes. Each point represents a single somatic mutation. Point color codes for protein domain. Total mutation numbers are displayed in the plot captions._

@@

```{r fig-genetic-mds, fig.width = figur::convert(rep_figs$genetic_mds, to = 'in')$w, fig.height = figur::convert(rep_figs$genetic_mds, to = 'in')$h, fig.cap = 'Co-occurrence of the most common genetic alterations in urothelial cancer.'}

rep_figs$genetic_mds$plot

@@

__Figure \@ref(fig:fig-genetic-mds). Co-occurrence of the most common genetic alterations in urothelial cancer.__

_Co-occurrence of the most common genetic alterations in urothelial cancers (at least 5% of cancer samples) was investigated in the `r globals$cohort_labs[c("genie", "msk", "tcga")] %>% collapse_and` cohorts by two-dimensional MDS (multi-dimensional scaling) of pairwise Jaccard's distances._ 
_MDS results are visualized in scatter plots with single genetic alterations depicted as points. Point color codes for alteration type (somatic mutation, deletion, amplification), point size represents frequency of the alteration in the data set. Note that genetic features displayed in the plots close to each other are expected to co-occur in a high fraction of cancer samples. Total numbers of samples are indicated in the plot captions._

@@

```{r fig-genetic-cont-hm, fig.width = figur::convert(rep_figs$genetic_cont_hm, to = 'in')$w, fig.height = figur::convert(rep_figs$genetic_cont_hm, to = 'in')$h, fig.cap = 'Co-occurrence of selected genetic alterations in urothelial cancer.'}

rep_figs$genetic_cont_hm$plot

@@

__Figure \@ref(fig:fig-genetic-cont-hm). Co-occurrence of selected genetic alterations in urothelial cancer.__

_Co-occurrence of the most common genetic alterations in urothelial cancers (at least 5% of cancer samples) was investigated in the `r globals$cohort_labs[c("genie", "msk", "tcga")] %>% collapse_and` cohorts by pairwise Jaccard's similarity coefficients J. Statistical significance of co-occurrence, i.e. $J > 0$, was assessed by false discovery rate-adjusted bootstrap tests._ 
_Frequencies of samples with and without genetic alterations are visualized as heat maps of contingency tables for selected pairs of genetic features. Sample percentages and counts within the total sample sets are indicated in the heat map tiles. J and p values are presented in the plot captions._

```{r fig-fgfr-mds, fig.width = figur::convert(rep_figs$fgfr_mds, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr_mds, to = 'in')$h, fig.cap = 'Co-occurrence of genetic alterations of FGF- and FGFR-coding genes in urothelial cancer.'}

rep_figs$fgfr_mds$plot

__Figure \@ref(fig:fig-fgfr-mds). Co-occurrence of genetic alterations of FGF- and FGFR-coding genes in urothelial cancer.__
